---
title: "Do Reddit /r/depression Authors who Pen Suicidal Ideation Leave Reddit at Higher Rates?"
author: "Lacey Etzkorn"
date: "October 18, 2016"
output: 
      html_document:
            toc: TRUE
---

```{r setup, include=FALSE}
require(ggplot2); require(dplyr); require(reshape2)
options(width = 90)
options(digits = 4)
load("June_Authors_Combined.RData")
```

# Introduction

The problem of suicide plagues social media and online communities devoted to mental health. [add context] 

Social media has also learned to fight back. [add context]

Given the delicate relationship between social media and suicide, it is of utmost importance that community moderators study the impacts of their policies on their community's risk of suicide. There are numerous stories of how seemingly benign changes in community moderation policies have exacerbated user's poor mental health status. For example, [give example].

Currently, moderators do not have data-intensive tools to evaluate the impact of policies on suicide risk. [Do they?]. This is due, in part, to the lack of data available to moderators. In most cases, it is impossible for moderators to observe actual outcomes of suicide, especially when users are anonymous. Surrogate outcomes for suicide are currently in development. For example, Munmun de Choudhury uses [linguistic machine learning methods] to identify and predict user-specific shifts from discussing poor mental health to discussing suicide ideation. However, detecting suicidal ideation in user posts has limited utility when the desired outcome is a rate of suicide.

In this paper, we propose a more direct method for estimating rates of suicide from social media data. [Do we actually do this?] We do this by estimating the rate at which users who admit to having suicidal ideation (SI) stop posting on Reddit. We then compare the drop-out rates between SI-positive and SI-negative users. Though our method is thoroughly problematic in its current state, we believe that, with a proper amount of statistical development and validation, it has the potential to become a decent measure of suicide rates in online mental health communities.

# Methods

[give methods overview]

The objective of our project is to quantify the risk of suicide for /r/depression authors. However, this objective is impossible for a number of reasons.

Another potential es

Given that the desired objective is impossible, we set out to estimate a reasonable surrogate parameter that will still be useful to online community moderators. Hence, we define the following parameter.

The difference in rate of dropouts 

The goal of our analysis is to estimate the difference in drop out rates between Reddit /r/depression authors who express suicidal ideation and /r/depression authors who don't

### Data

Our target population is Reddit authors who post in the subreddit /r/depression in the month of June. Hence, we scrape all available (as of October 10, 2016) comment and post data from /r/depression for the month of June using the python package subredditarchive. The month June was chosen to ensure that we have at least three months of follow-up time for each author. An earlier month wasn't chosen because Reddit censors author comment and post history to the most recent 1,000 items each. Henceforth we will refer to comments and posts collectively as posts.

We then identified whether the text from each post contained evidence of suicidal ideation (SI) according to our classification algorithm. Details of this classification process are described in a subsequent section. Then, we produced one list of unique author ids associated with any SI evidence as well as one equal-sized list of unique authors without any SI evidence. We then scraped the post history for both lists of authors and identified posts with SI evidence.

### Estimation of Drop-Out Rates

Our goal is to estimate the expected rate at which SI-positive and SI-negative authors stop using Reddit. Ultimately, we will compare our estimated rates between these two groups to determine if there is some association between suicidal ideation and risk of dropping-out. 

First, we estimate the author-specific average time between posts $mu_i$, where $i$ indexes the authors.
We assume that the post histories of each author are a Poisson process, and so observed times between posts are exponentially distributed with mean $mu_i$. For each $\mu_i$, we can specify the prior distribution $1/\mu_i \sim gamma(1,1)$. Hence the posterior for $\mu_i$ is $1/\mu_i|n,\bar{\epsilon} \sim gamma(1 + n, 1 + n\cdot\bar{\epsilon})$, where n is the author's number of intervals between posts, and $\bar{\epsilon}$ is the user-specific average time between posts.

From each $mu_i$, we are able to calculate the probability that the author has stopped using Reddit.
Let $\rho$ be the prior probability that any author drops out of Reddit. Let $t_i$ be the time between data collection and the most recent post for user $i$. This variable can be thought of the "censoring time". Then, the posterior probability that a user quits given their true time between posts is:
$$ p_i|\mu_i   =   \frac{\rho} {e^{-\mu_i\cdot t_i} \cdot (1-\rho) + \rho} $$ 
With this expression, we can transoform the monte-carlo samples from the posterior of $\mu_i$ to monte-carlo samples from the posterior of $p_i$.

Finally, we can calculate the posterior distribution of the expected number of Reddit drop-outs. If $Q_i$ is the indicator that user $i$ has dropped out, then $E(\sum Q_i \Big | p_1, p_2 ...) = E(\sum p_i \Big | p_1, p_2 ...)$.

Using this method, we generate monte-carlo samples from the posterior distribution for the expected number of drop-outs among SI-positive authors and SI-negative authors each. We can then compare these posteriors to get a sense of the number of excess Reddit drop-outs associated with suicidal ideation.

### SI Classification

Our goal for our SI classification algorithm was to produce a set of comments that almost certainly contained only true SI comments. Hence, we set out to maximize specificity while throwing sensitivity out the window. While we recognize that could be many algorithms better suited for our final goal of estimating the incidence of suicide in our population, we were fine to just make something that worked for now.

[Definition of Suicidal Ideation]

We began by reading through a sample of 100 randomly selected comments from various mental health subreddits to get a sense of how people expressed suicidal ideation on Reddit in general. From each of the SI comments from this 100, we extracted the smallest phrase(s) that contained evidence of SI. Examples of these phrase bits include: 

* "I think about suicide"
* "I just don't want to live""
* "I have struggled with depression and suicidal thoughts"
* "my suicide plan"
* "I still feel like blowing my brains out"

We then categoized these phrases into 12 linguistically distinct modes of expressing SI, adding extra where we felt appropriate:

Intention to, or thoughts of suicide/dying:
* [I am] [intending to] [suicide verb]
* [I am] [thinking about] [suicide noun]
* I [intend to] [suicide verb]
* I [suicide verb]
* [I am] [suicide adjective]
* Why not [suicide verb]
* I [think about] [suicide noun]
* I [think]
* me [suicide verb]
* me [suicide noun]

Oppose life:
* [I] [oppose] [live noun]
* [I] [am opposed to] [living]

For each of these modes of SI expression, we generated all possible combinations of words for each category to get all possible phrases for that mode. For example, to generate all possible phrases for  "I [intend to] [suicide verb]", we combined the "intend to" verbs 

      want to, were to, would like to, wish to, decided to, decide to, am not afraid to, am not scared to, have to, had to, try to, tried to, would like to, can, will, must, might, should, could, intend to, intended to, have come to, 've come to, ‘d like to, ‘d, ‘ll

with the suicide verbs

      die, commit suicide, kill myself, off myself, end my life, no longer be here, take some pills, leave this world, end it, blow my brains out, kicked the bucket, kick the bucket, cut my throat, hang myself

After all phrases for all 12 modes of SI expression have been generated, we remove all punctuation and spacing from both the post text and the phrase text. We combine all phrases into a regular expression using the or "|" operator, and then use the function "grepl" to test whether each post contains any phrase.

# Results

### June /r/depression

For the month of June, /r/depression contained 22,041 posts, representing contributions from 7,572 authors. We identified 547 SI-positive posts from 492 unique users. 

In a sample of 100 SI-positive posts, [n] did not have true SI evidence. In a sample of 100 SI-negative posts, [n] did have SI evidence. 

### User Follow-Up

```{r , echo=FALSE, message=FALSE, warning=FALSE}
user.data = 
      author.summary %>%
      group_by(author) %>%
      filter(newest == max(newest)) %>%
      ungroup %>%
      group_by(si.author)%>%
      transmute(author, rank_newest = rank(newest)) %>%
      ungroup %>%
      select(-si.author) %>%
      merge(user.data, by = "author", all.y=T)
user.data$point.label = 
      ifelse(user.data$class, "SI",
      ifelse(user.data$r.depression, "/r/depression","other")) %>%
      ordered(labels = c("other", "/r/depression", "SI"))
user.data$si.group = 
      ifelse(user.data$si.author, "SI-Positive in June /r/depression",
             "SI-Negative in June /r/depression")
### Make plot
ggplot(data = subset(user.data, date > jan1)) +
geom_point(aes(y=rank_newest,x=date, 
               color=point.label, alpha=point.label, 
               size=point.label)) + 
ggtitle("Distribution of Post Dates across Authors\nbetween Jan. and Oct.") +
ylab("Author Rank by Newest Post") + xlab("Date") +
geom_vline(aes(xintercept = as.numeric(july1)), alpha=0.3) + 
geom_vline(aes(xintercept = as.numeric(june1)), alpha=0.3) + 
scale_colour_manual(values = c("grey","darkblue", "red")) + 
scale_size_manual(values = c(0.1,0.1,0.3)) + 
scale_alpha_manual(values = c(0.1,0.1,1)) + 
theme_bw() + facet_wrap("si.group")
```

After scraping the post history for each of the SI-positive users, we find no post or comment data for 1 author, and we find 16 authors whose full history was not available back to April 2016. We are left with 475 authors. 

The post histories for our 475 authors are depicted above. Each y value represents a unique author id, sorted by the time of their most recent post. Each x value represents the date of a post. Hence, we can read each horizontal line on this plot as the post history of one author. Red points represent SI-positive posts. Blue points represent posts in the /r/depression subreddit. Grey posts represent posts elsewhere on Reddit. There a number of important features in this plot. First, we note that those authors who posted more recently tended to post more often, which is expected. Second we see that many authors made multiple SI-positive posts. Third, we see that nearly 170 of our 475 authors had their last post in June, which was the period that we used to sample from. Fourth, of these 170 authors, many did not post frequently before dropping-off Reddit. However, a hand-full of them had posted frequently, and their drop-put is therefore suspicious.

### Drop-Out Rate Estimates

The histogram below represents the posterior distribution for the rate of drop-outs for authors whose June /r/depression comments were SI-positive.

```{r, echo=FALSE, warning=FALSE}
### Load Data
      load(file = "Data/SI_Author_Sample/Author_Summary.RData")
      user.summary = 
            filter(user.summary, recent < 150) %>%
            mutate(t = recent)
      n = user.summary$n
      eps = user.summary$eps
      eps[is.na(eps)] = 0 
      t = user.summary$t
      N = nrow(user.summary)
### Simulate Mu given Data
      alpha = 1; beta = 1
      mu = replicate(n = 10000, rgamma(N, shape = alpha + n, rate = beta + n*eps))
### Now get overall credible interval for sum(Q)
      prior.p = 0.000001
      make.p = function(mu,p) p / (exp(-mu*t)*(1-p) + p)
      post.E = colSums(apply(mu, 2, make.p, p=prior.p))/N * 100
      hist(post.E, main = "Posterior Distribution for\nExpected Reddit Drop-Outs per 100 SI-Positive Authors\n(Prior = 10e-4)",
           xlab="", ylab="")
      ci = quantile(post.E, probs = c(0.025, 0.975))
```

Our 95% credible interval is (`r ci`), meaning there is a 95% probability that the expected rate of drop-out is between `r ci[1]` and `r ci[2]`.

### Sensitivity to Priors

We found that our conclusions were highly sensitive to our specified prior probability of drop-outs.

```{r, echo=FALSE, warning=FALSE}
### Calculate P(Q) from Mu and data
      prior.p = c(0.5, 10^-(1:4))
      post_p = sapply(prior.p, FUN = function(x) apply(mu, 2, make.p, p=x) %>% rowMeans)
      colnames(post_p) = c("p.5", paste0("p10.", 1:4))
      user.summary2 = 
            cbind(user.summary, post_p) %>%
            select(-recent) %>%
            melt(id.vars = c("author", "n", "eps", "t"),
                 value.name = "post.p", variable.name = "prior.p")
      ggplot(user.summary2) + 
            geom_point(aes(t, eps, color=post.p), size=0.5, alpha=0.5) + 
            facet_wrap("prior.p") + 
            theme_bw() + 
            xlab("Average Days between Posts") +
            ylab("Days since Most Recent Post")
```

# Discussion

Online communities present a unique platform where people can address their own mental health concerns. 

A online community such as the depression subreddit provides a unique opportunity for researchers to glimpse into the thoughts of people with depression outside of a clinical setting. The depression subreddit is a public community where users do not need to present any identifiable information. 

### Limitations

For this method, we only incorporate user specific information as the time between posts. However, a number of other factors might give us information about whether a user actually attempts suicide. Sentiment might matter. Recent SI posts might matter. Actual suicide notes might matter.

We do not differentiate between suicidal ideation and suicide notes. 

We also cannot observe unsuccessful suicide attempts.


# Works Cited

* Discovering Shifts to Suicidal Ideation from Mental Health Content in Social Media (2016) http://www.munmund.net/pubs/chi16_SuicidalIdeation.pdf 

* Social Media as a Measurement Tool of Depression in Populations (2013) http://www.munmund.net/pubs/websci_13.pdf

* Predicting Postpartum Changes in Emotion and Behavior via Social Media (2013) http://www.munmund.net/pubs/chi_13.pdf

* Social Media Embraces Suicide Prevention CMAJ August 11, 2015 vol. 187 no. 11 First published June 29, 2015, doi: 10.1503/cmaj.109-5104 http://www.cmaj.ca/content/187/11/E333.short 


